\chapter{Proposed Work}

The study aims to explore both grid-based algorithms and more complex network models, focusing on understanding disease spread dynamics and improving the traceability of infection sources.

\section{Grid-Based Models}

Grid-based models are fundamental in epidemiological studies due to their simplicity and structured nature. These models represent the population as nodes in a grid, where each node can be in one of several states: susceptible, infected, or recovered. The spread of disease is simulated by allowing the infection to propagate from one node to its neighboring nodes.

The infection probabilities on neighbors in a grid-based model were discussed in detail in the Mathematical Modeling chapter along with the interaction rules (Section \ref{section:infection_probabilities}).

\label{section:boundary_conditions}
Moreover, boundary conditions are crucial in grid-based models as they influence the spread of infection, especially at the edges of the grid. There are two primary types of boundary conditions:
\begin{itemize}
    \item \textbf{Fixed Boundary:} In a fixed boundary grid, nodes at the edges have fewer neighbors compared to nodes in the interior. This can create edge effects where the infection dynamics at the boundaries differ significantly from those in the central region.
    \item \textbf{Wrap-Around Boundary (Toroidal Grid):} In a wrap-around grid, the edges are connected in such a way that the grid forms a torus. This means that nodes at the edges are connected to nodes on the opposite edge, eliminating boundary effects and creating a uniform infection spread throughout the grid.
\end{itemize}

The structure and configuration of the grid, along with the interaction rules, determine how the infection spreads. In our project, the grid is configured as a square lattice where each node has up to four neighbors (north, south, east, west).

Note that grid-based models provide a foundational approach to studying the spread of infectious diseases. By carefully considering infection probabilities, boundary conditions, and interaction rules, these models can offer valuable insights into the dynamics of disease transmission and the effectiveness of various intervention strategies.

\subsection{Algorithm Selection}
Three primary algorithms were selected for their distinct approaches and potential effectiveness in source identification within grid-based models:

\begin{enumerate}
    \item Jordan Center Identification using BFS
    \item Euclidean Distance Minimization
    \item Center of Mass Identification
\end{enumerate}

\subsection{Performance Metrics}
The performance of the selected algorithms is evaluated based on the following criteria:
\begin{enumerate}
    \item Accuracy
    \item Complexity
    \item Handling Boundaries
\end{enumerate}

\section{Complex Network Models}
The exploration of complex network models is essential for a comprehensive understanding of disease spread dynamics in real-world scenarios. Unlike grid-based models, complex networks exhibit diverse structures and connectivity patterns, which can significantly influence the spread of infections.

\subsection{Network Types}

\subsubsection{Erdős–Rényi Graphs}
Erdős–Rényi graphs are characterized by having a fixed number of nodes where each pair of nodes is connected with a fixed probability. They are useful for modeling random connections in a network, providing a basis for understanding the impact of randomness on disease spread.

\subsubsection{Random Regular Graphs}
Random Regular graphs consist of nodes each with the same number of connections (degree), ensuring uniformity in node connectivity. They help in examining disease spread in networks with uniform connectivity, highlighting the effect of equal distribution of contacts.

\subsubsection{Random Geometric Graphs}
Random Geometric graphs are constructed by placing nodes randomly in a geometric space and connecting nodes that are within a certain distance from each other. They simulate spatial networks, making them suitable for studying geographic constraints on the spread of infections.

\subsection{Algorithm Selection}
For the analysis of infection spread on complex networks, several algorithms were selected based on their effectiveness in identifying the source of infection:

\begin{enumerate}
    \item Jordan Center Identification
    \item Eccentricity and Closeness Centrality
    \item Betweenness Centrality
    \item Eigenvector Centrality
    \item Closeness Centrality
\end{enumerate}

\subsection{Performance Metrics}
The performance of the algorithms on complex networks is measured based on the following criteria:
\begin{enumerate}
    \item Accuracy: The precision of the algorithm in correctly identifying the source of infection.
    \item Scalability: The ability of the algorithm to handle increasing network sizes effectively.
\end{enumerate}


\section{Stochastic Approach: Stochastic Differential Equations (SDEs)}
We employed stochastic differential equations to introduce randomness into the infection spread models. They allow for the incorporation of stochastic variations and uncertainties in the transmission process. The Gillespie Algorithm will be employed to simulate the stochastic models. This algorithm is particularly suitable for simulating the time evolution of systems with random events, such as the spread of infectious diseases. We'll then compare the stochastic approach to the deterministic approach.

\section{Conclusion}
This chapter has outlined the proposed work, detailing the objectives and expected outcomes of the research. The focus is on comparing grid-based algorithms and complex network models to enhance the understanding of disease spread dynamics.

In the next chapter, we will detail the design and implementation of the models and algorithms used in this study.